---
title: "Tidy Intro Talk"
author: "Michael Love"
format: html
editor: visual
---

## Data objects keep data organized

```{r}
#| message: false
library(SummarizedExperiment)
genes <- DataFrame(id = paste0("g",1:4), 
                   symbol = c("ABC","DEF","GHI","JKL"))
genes
```

```{r}
samples <- DataFrame(sample = paste0("s",1:4),
                     condition = c("x","y","x","y"),
                     sex = c("m","m","f","f"))
samples
```

```{r}
counts <- matrix(rpois(16, lambda=100), ncol=4)
rownames(counts) <- genes$id
colnames(counts) <- samples$sample
counts
```

```{r}
se <- SummarizedExperiment(list(counts = counts),
                           rowData=genes, 
                           colData=samples,
                           metadata=list(organism="Hsapiens"))
se
```

## Objects help avoid common bookkeeping errors

Reordering samples (columns) is a global operation:

```{r}
se2 <- se[,c(1,3,2,4)]
assay(se2, "counts")
colData(se2)
```

Assignment that might result in sample swap results in an error:

```{r}
#| error: true
assay(se2) <- counts
```

Other such checks include comparison of genomic ranges across
different genome builds: will result in an error.

## However, hard to get started for new users

```{r}
methods(class=class(se))
```

## And yet, many beginners know dplyr and ggplot2

```{r}
#| message: false
library(dplyr)
samples |> 
  as_tibble() |> 
  filter(condition == "x")
```

## tidyomics enables tidy verbs for omics objects

```{r}
#| message: false
library(tidySummarizedExperiment)
se
```

What does this mean "*SummarizedExperiment-tibble abstraction*"?

...put a drawing...

## This is still a standard Bioconductor object

```{r}
se |> 
  filter(condition == "x")
```

```{r}
se_sub <- se |> 
  filter(condition == "x")
class(se_sub)
colData(se_sub)
```

## Facilitates quick exploration and plotting

```{r ggplot}
#| message: false
library(ggplot2)
se |> 
  filter(.feature %in% c("g1","g2")) |> 
  ggplot(aes(condition, counts, color=sex, group=sex)) + 
  geom_point(size=2) + 
  geom_line() +
  facet_wrap(~.feature) +
  ylim(0,200) + 
  theme(text = element_text(size = 16))
```
